2.1: Data 101 and Descriptive Statistics

### 2019-09-17 static/slides/07-slides.Rmd
Tweaks to slides 7

# Data 101

@@

- .hl[Data] are information with context
- .hl[Individuals] are the entities described by a set of data 
    - e.g. persons, households, firms, countries 
- .hl[Variables] are particular characteristics about an individual
    - e.g. age, income, profits, population, GDP, marital status, type of legal institutions
- .hl[Observations] or .hl[cases] are the separate individuals described by a collection of variables
    - e.g. for one individual, we have their age, sex, income, education, etc.  
    - individuals and observations are *not necessarily* the same: 
        - e.g. we can have separate observations on the same individual over time 
]

@@

- .hl[Categorial data] place an individual into one of several possible *categories*
    - e.g. sex, season, political party
    - may be responses to survey questions
    - can be quantitative (e.g. age, zip code)

@@

- See today's class notes for more about the summation operator, $\displaystyle\Sigma$, it'll come up again!

@@

.pull-left[
.font90[
1. Minimum value
2. 25<sup>th</sup> percentile $(Q_1$, the median of the first 50% of data)
3. 50<sup>th</sup> percentile (median, $Q_2)$
4. 25<sup>th</sup> percentile $(Q_3$, the median of the last 50% of data)
5. Maximum value
]
]
.pull-right[

@@

- Height of box: $Q_1$ to $Q_3$ (known as .onfire[interquartile range (IQR)], middle 50% of data)
- Line inside box: median (50<sup>th</sup> percentile)
- "Whiskers" identify data within $1.5 \times IQR$
- Points *beyond* whiskers are a common definition of **outliers**
]
.pull-right[

@@

```{r, echo=T}
quizzes_new %>% summary(quiz_1)

quizzes_new %>% summary(quiz_2)
```
]

@@

- So I wrote my own `R` function that makes nice summary tables that uses `dplyr` and `tidyr`

@@

```{r, echo=T}
# loads .R files to full functions from; make sure it's in working directory/Project!
source("../files/summaries.R")

# let's summarize "hwy" and "cty" from mpg dataset
mpg %>%
  summary_table(hwy, cty)
```

@@

- Every observation .onfire[deviates] from the mean of the data: 
$$deviation_i = x_i-\mu	$$

- There are as many deviations as there are data points $(n)$
- We can measure the *average* or .shout[standard deviation] of a variable from its mean

@@

- The .shout[population variance `\\((\sigma^2)\\)`] of a *population* distribution measures the average of the *squared* deviations from the *population* mean

@@

- The .shout[sample variance `\\((s^2)\\)`] of a *sample* distribution measures the average of the *squared* deviations from the *sample* mean

$$\sigma^2 = \frac{1}{n-1}\displaystyle\sum^n_{i=1} (x_i-\mu)^2$$

@@

$$s=\sqrt{s^2}=\sqrt{\frac{1}{n-1}\displaystyle\sum^n_{i=1} (x_i-\mu)^2	}$$

>>>>

# Data 101 I

@@

- .hl[Data] are information with context

- .hl[Individuals] are the entities described by a set of data 
    - e.g. persons, households, firms, countries

# Data 101 I

- .hl[Variables] are particular characteristics about an individual
    - e.g. age, income, profits, population, GDP, marital status, type of legal institutions

- .hl[Observations] or .hl[cases] are the separate individuals described by a collection of variables
    - e.g. for one individual, we have their age, sex, income, education, etc.

- individuals and observations are *not necessarily* the same: 
    - e.g. we can have separate observations on the same individual over time 
]

@@

- .hl[Categorical data] place an individual into one of several possible *categories*
    - e.g. sex, season, political party
    - may be responses to survey questions
    - can be quantitative (e.g. age, zip code)

@@

- See today's class notes for more about the .onfire[summation operator, `\\(\displaystyle\Sigma\\)`], it'll come up again!

@@

.pull-left[
.font90[
1. Minimum value
2. 25<sup>th</sup> percentile $(Q_1$, the median of the first 50% of data)
3. 50<sup>th</sup> percentile (median, $Q_2)$
4. 25<sup>th</sup> percentile $(Q_3$, the median of the last 50% of data)
5. Maximum value
]
]

@@

- **Height of box**: $Q_1$ to $Q_3$ (known as .onfire[interquartile range (IQR)], middle 50% of data)

- **Line inside box**: median (50<sup>th</sup> percentile)

- **"Whiskers"** identify data within $1.5 \times IQR$

- Points *beyond* whiskers are **outliers**
    - common definition: $Outlier >1.5 \times IQR$
]
.pull-right[

@@

```{r, echo=T}
quizzes_new %>% summary()
```
]

@@

- So I wrote my own `R function` that makes nice summary tables that uses `dplyr` and `tidyr`

@@

```{r, echo=T}
# loads .R files to full functions from; make sure it's in YOUR working directory/Project!
source("../files/summaries.R") # MY path (this website's R Project)

# let's summarize variables "hwy" and "cty" from *mpg* dataset
mpg %>%
  summary_table(hwy, cty)
```

@@

- Every observation $i$ .onfire[deviates] from the mean of the data: 
$$deviation_i = x_i-\mu	$$

- There are as many deviations as there are data points $(n)$

- We can measure the *average* or .shout[standard deviation] of a variable from its mean

- Before we get there...

@@

- The .shout[population variance `\\((\sigma^2)\\)`] of a *population* distribution measures the average of the *squared* deviations from the *population* mean $(\mu)$

@@

- The .shout[sample variance `\\((s^2)\\)`] of a *sample* distribution measures the average of the *squared* deviations from the *sample* mean $(\bar{x})$

$$\sigma^2 = \frac{1}{n-1}\displaystyle\sum^n_{i=1} (x_i-\bar{x})^2$$

@@

$$s=\sqrt{s^2}=\sqrt{\frac{1}{n-1}\displaystyle\sum^n_{i=1} (x_i-\bar{x})^2	}$$
